---
title: 'Reproducible Research: Peer Assessment 1'
author: William L. Kennedy
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading and preprocessing the data

```{r}
# Package dependencies (uncomment and install if necessary)

# install.packages("data.table")
# install.packages("lubridate")
library(data.table)
library(lubridate)

# set the working directory (replace path accordingly)
setwd("/Users/adakemia/Documents/Academic/Coursera/DataScienceSpecialization/05ReproducibleResearch/Projects/RepData_PeerAssessment1")

# List and check for necessary files
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
zfile <- "activity.zip"
file <- "activity.csv"

if (!file.exists(file)) {
        if (!file.exists(zfile)) {
                method <- switch(Sys.info()[['sysname']],
                                 "Windows" = "internal",
                                 "Darwin" = "curl",
                                 "Linux" = "wget",
                                 "auto")
                download.file(url, destfile=zfile, method)
                }
        unzip(zfile, exdir="./")
}

# Read files into data.table object
data <- fread(file, sep=",", stringsAsFactors=F, header=T, 
              na.strings="NA", colClasses=c("numeric","character","character"))
# Check structure of data
str(data)

# Process data

# Set date formatting (lubridate package)
data[,date := ymd(date)]

# Data quality checks

# Check for missingness
colSums(is.na(data))
# Percent missing
sum(is.na(data$steps)) / nrow(data) * 100

# Check timeseries
data[,unique(date)]
plot(data$interval[11:14], type="l")
data[11:14]

# Fix hour/min anomaly


```


## What is mean total number of steps taken per day?  
First, we can take a wide view via a histogram. We can see several things from the histogram:  
* The shape is relatively normal  
* Due to normality, we would expect mean and median to be fairly close  
* We would expect the mean and median to fall roughly just over 10,000 steps  

---

```{r}
hist(data[,sum(steps), by=date]$V1, breaks=8, 
     main="Histogram of Mean Steps Per Day",
     xlab="mean steps per day")
```
  
Next, we can compare to the actual mean and median for the entire period. As expected, the mean and median are very close and just over 10,000 steps.  


```{r}
data[, .(sum = sum(steps)), by=date][,.(median = median(sum, na.rm=T),
                                        mean = mean(sum, na.rm=T))]
```

## What is the average daily activity pattern?

```{r}
ts <- data[, .(mean = round( mean(steps, na.rm=T), 2)), by=interval]

plot(ts$mean ~ ts$interval, type="l", main="Mean Steps Per Daily Interval",
     xlab="interval", ylab="mean steps")

ts[which.max(mean)]

```
## Imputing missing values

```{r}
# create a second imputed data set
data_imp <- data

ts2 <- data_imp[, .(median = median(steps, na.rm=T),
                    mean = mean(steps, na.rm=T)), 
                by=interval]
# Compare mean and median by interval 
plot(ts2$mean ~ ts2$interval, type="l", main="Mean Steps Per Daily Interval",
     xlab="interval", ylab="mean steps")
lines(ts2$median ~ ts2$interval, col="red")

data[interval == 1000]
data_imp[, .(median = median(steps, na.rm=T),
                    mean = mean(steps, na.rm=T)), 
                by=list(interval, wday(date))][interval==900]
ts2[interval == 900]
data[interval == 900]
```

## Are there differences in activity patterns between weekdays and weekends?
